Simple repression draft

### 2020-05-29 doc/appendix_02_bursty.tex
MRM going through RP comments on bursty dist. derivation

The objective of this section is to write down the steady-state mRNA
distribution for model 5 in Figure~\ref{fig2:constit_cartoons}. Our claim is
that this model is rich enough that it can capture the expression pattern of
bacteria unregulated promoters regulated by the $\sigma_{70}$ sigma factor.
Figure~\ref{figS1:bursty_one_state} shows two different schematic
representations of the model. Figure~\ref{figS1:bursty_one_state}(A) shows the
promoter cartoon model with burst initiation rate $k_i$, mRNA degradation rate
$\gamma$, and mean burst size $b$. For our derivation of the chemical master
equation we will focus more on Figure~\ref{figS1:bursty_one_state}(B). This
representation is intended to highlight that bursty gene expression allows
transitions between mRNA count $m$ and $m'$ even with $m - m' > 1$.

\begin{figure}[h!]
\centering
\includegraphics{../figures/si/figS0X_bursty_states.pdf}
\caption{
\textbf{Bursty transcription for unregulated promoter.}
(A) Schematic of the one-state bursty transcription model. Rate $k_i$ is the
bursty initiation rate, $\gamma$ is the mRNA degradation rate, and $b$ is the
mean burst size. (B) Schematic depiction of the mRNA count state transitions.
The model in (A) allows for transitions of $> 1$ mRNA counts with probability
$G_{m-m'}$.}
\label{figS1:bursty_one_state}
\end{figure}

To derive the master equation we begin by thinking the possible state
transitions to ``enter'' state $m$. There are two possible paths to jump from an
mRNA count $m' \neq m$ to a state $m$ in a small time window $\Delta t$:
\begin{enumerate}
        \item By degradation of a single mRNA, jumping from $m+1$ to $m$.
        \item By producing $m-m'$ mRNA for $m' \in \{0, 1, \ldots, m-1\}$.
\end{enumerate}
For the ``exit'' states from $m$ into $m' \neq m$ during a small time window
$\Delta t$ we also have two possibilities:
\begin{enumerate}
        \item By degradation of a single mRNA, jumping from $m$ to $m-1$.
        \item By producing $m'-m$ mRNA for $m'-m \in \{1, 2, \ldots\}$.
\end{enumerate}
This implies that the probability of having $m$ mRNA at time $t + \Delta t$ can
be written as
\begin{equation}
\begin{split}
p(m, t + \Delta t) = &p(m, t)
+ \overbrace{\gamma \Delta t (m + 1) p(m + 1, t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma \Delta t m p(m, t)}^{m \rightarrow m - 1} \\
&+ \overbrace{k_i \Delta t \sum_{m'=0}^{m-1} G_{m-m'} p(m', t)}^
{m'\in \{0, 1, \ldots m-1\} \rightarrow m}
- \overbrace{k_i \Delta t \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t)}^
{m \rightarrow m'\in \{m+1, m+2, \ldots\}},
\end{split}
\label{eq:si_master_deltat}
\end{equation}
where we indicate $G_{m'-m}$ as the probability of having a burst of size
$m'-m$. We suggestively use the letter $G$ as we will assume that these bursts
sizes are geometrically distributed with parameter $\theta$. This is written as
\begin{equation}
G_{k} = \theta (1 - \theta)^k\; \text{for } k \in \{0, 1, 2, \ldots \}.
\end{equation}
What this implies is that for a geometrically distributed burst size we have a
mean burst size $b$ of the form
\begin{equation}
b \equiv \left\langle m' - m \right\rangle 
= \sum_{k=0}^\infty k \theta (1 - \theta)^k = {1 - \theta \over \theta}.
\end{equation}

To clean up Equation~\ref{eq:si_master_deltat} we can send the first term on the
right hand side to the left, and divide both sides by $\Delta t$. Upon taking
the limit where $\Delta t \rightarrow 0$ we can write
\begin{equation}
{d \over dt}p(m, t) = (m + 1) \gamma p(m + 1, t)
- m \gamma p(m, t)
+ k_i \sum_{m'=0}^{m-1} G_{m-m'} p(m', t) 
- k_i \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t).
\end{equation}
Furthermore, given that the timescale for this equation is set by the mRNA
degradation rate $\gamma$ we can divide both sides by this rate, obtaining
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau),
\label{eq:si_master_ode}
\end{equation}
where we defined $\tau \equiv t \times \gamma$, and $\lambda \equiv k_i/\gamma$.
The last term in Eq.~\ref{eq:si_master_ode} sums all burst sizes except for a 
burst of size zero. We can re-index the sum to include this term, obtaining
\begin{equation}
\lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau) = \lambda p(m, t) \left[
\underbrace{\sum_{m'=\textcolor{red}{m}}^{\infty}G_{m'-m}}
_{\text{re-index sum to include burst size zero}} -
\underbrace{G_0}_{\text{subtract extra added term}}\right].
\end{equation}
Given the normalization of the geometric distribution, adding the probability of
all possible burst sizes -- including size zero since we re-indexed the sum --
allows us to write
\begin{equation}
\sum_{m'=m}^{\infty}G_{m'-m} - G_0 = 1 - G_0.
\end{equation}
Substituting this into Eq.~\ref{eq:si_master_ode} results in
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda p(m, \tau) \left[ 1 - G_0 \right].
\label{eq:si_master_ode_2}
\end{equation}
To finally get at a more compact version of the equation notice that the third
term in Eq.~\ref{eq:si_master_ode_2} includes burst from size $m'-m = 1$ to size
$m' - m = m$. We can include the term $p(m, t) G_0$ in the sum which allows
bursts of size $m' - m = 0$. This results in our final form for the chemical
master equation
\begin{align}
{d \over d\tau}p(m, \tau) = 
(m + 1) p(m+1, \tau)
- m p(m, \tau) - 
\lambda p(m, \tau)
+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime} p(m^\prime, \tau).
\label{eq:si_master_unreg}
\end{align}

In order to solve Eq.~\ref{eq:si_master_unreg} we will use the generating
function method \mrm{reference?}. The probability generating function is defined
as
\begin{align}
F(z,t) = \sum_{m=0}^\infty z^m p(m,t),
\end{align}
where $z$ is just a dummy variable that will help us later on to obtain the
moments of the distribution. Applying $F(z, t)$ on both sides of 
Eq.~\ref{eq:si_master_unreg} results in
\begin{equation}
\sum_m z^m {d \over d\tau} p(m, \tau) = 
\sum_m z^m \left[ 
- m p(m, \tau) 
+ (m + 1) p(m + 1, \tau) 
+ \lambda \sum_{m' = 0}^m G_{m-m'} p(m', \tau) - \lambda p(m, \tau)
\right],
\end{equation}
where we use $\sum_m \equiv \sum_{m=0}^\infty$. We can distribute the sum and
use the definition of $F(z, t)$ to obtain
\begin{equation}
{d F(z, \tau) \over d\tau} =
- \sum_m z^m m p(m, \tau) +
+ \sum_m z^m (m + 1) p(m + 1, \tau)
+ \lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau)
- \lambda F(z, \tau).
\label{eq:si_generating_01}
\end{equation}
We can make use of properties of the generating function to write everything in
terms of $F(z, \tau)$: the first term on the right hand side of
Eq.~\ref{eq:si_generating_01} can be rewritten as
\begin{align}
\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=
\sum_{m} z \frac{\partial z^{m}}{\partial z} p(m, \tau), \\
&=\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
&=z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
&= z {\partial F(z, \tau) \over \partial z}.
\end{align}
For the second term on the right hand side of Eq.~\ref{eq:si_generating_01} we
define $k \equiv m + 1$. This allows us to write
\begin{align}
\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=
\sum_{k=1}^{\infty} z^{k-1} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
&=\frac{\partial F(z)}{\partial z}.
\end{align}
The third term in Eq.~\ref{eq:si_generating_01} is the most trouble. The trick
is to reverse the default order of the sums as
\begin{equation}
\sum_{m=0}^{\infty} \sum_{m'=0}^{m} = \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty}.
\end{equation}
This results in
\begin{align}
\lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau) =
\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau),
\end{align}
where we also substituted the definition of the geometric distribution $G_{k} =
\theta (1 - \theta)^k$. Redistributing the sums we can write
\begin{align}
\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau) = 
\lambda \theta \sum_{n=0}^{\infty}(1-\theta)^{m'} P(m', \tau) 
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m}.
\label{eq:si_generating_02}
\end{align}
We know that $1 - \theta \in (0, 1)$, and that $z$ is a dummy variable that we
usually evaluate as $z=1$. With these two facts we can safely state that
$| z (1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in 
Eq.~\ref{eq:si_generating_02} as
\begin{align}
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &= 
\sum_{n=0}^{\infty} \left[z (1-\theta)\right]^{n + m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{n} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\left( {1 \over 1 - z (1 - \theta)} \right),
\end{align}
where we use the geometric series since, as stated before, $| z (1 - \theta) | <
1$. Putting these results together, the PDE for the generation function is
\begin{equation}
{\partial F \over \partial \tau} = 
{\partial F \over\partial z}
- z {\partial F \over \partial z} - \lambda F
+ \frac{\lambda\theta F}{1-z(1-\theta)}.
\end{equation}
Changing variables to $\xi=1-\theta$ and simplifying gives
\begin{align}
{\partial F \over \partial \tau} + (z - 1) {\partial F \over \partial z} = 
\frac{(z-1)\xi}{1-z\xi}\lambda F.
\label{eq:1state_unreg_015}
\end{align}

\subsubsection{Stead-state}

To get at the mRNA distribution at steady state we first must solve
Eq.~\ref{eq:1state_unreg_015} setting the time derivative to zero. At
steady-state, the PDE reduces to the ODE
\begin{align}
\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
\end{align}
which we can integrate as
\begin{align}
\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
\end{align}
The initial conditions for generating functions can be subtle and confusing. The
key fact follows from the definition
$F(z,t) = \sum_m z^m p(m,t)$,
normalization requires that
$F(z=1, t) = \sum_m p(m,t) = 1$.
\footnote{
Sometimes the generating function may be undefined \textit{at} $z=1$ but the
limit still holds. Also many authors change variables from $z$ to other
things, so one must keep track of how the normalization condition transforms.
}
Doing the integrals (and producing constant $c$) gives
\begin{align}
\ln F &= -\lambda \ln(1-\xi z) + c
\\
F &= \frac{c}{(1-\xi z)^\lambda}.
\end{align}
Only one choice for $c$ can satisfy initial conditions, producing
\begin{align}
F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
\end{align}
which is exactly the negative binomial's generating function. Now we may recover
the mRNA probability distribution of interest by differentiating the generating
function, or by consulting a standard reference\mmnote{maybe van Kampen, check},
from which we find that\mrm{I have a series of notes where I go through some 
of the derivatives to find the general pattern, arriving to this exact result 
by ``scratch and sniff'' without the black magic. Should I add it to this 
section?}
\begin{equation}
p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
        \theta^\lambda (1-\theta)^m.
\end{equation}
Note that the ratio of gamma functions is often expressed as a binomial
coefficient, but since $\lambda$ may be non-integer, this would be ill-defined.
Re-expressing this exclusively in our variables of interest, burst rate
$\lambda$ and mean burst size $b$, we have
\begin{equation}
p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
        \left(\frac{1}{1+b}\right)^\lambda
        \left(\frac{b}{1+b}\right)^m.
\label{eq:nbinom_deriv_final}
\end{equation}

>>>>

The objective of this section is to write down the steady-state mRNA
distribution for model 5 in Figure~\ref{fig2:constit_cartoons}. Our claim is
that this model is rich enough that it can capture the expression pattern of
bacteria constitutive promoters. Figure~\ref{figS1:bursty_one_state} shows two
different schematic representations of the model.
Figure~\ref{figS1:bursty_one_state}(A) shows the promoter cartoon model with
burst initiation rate $k_i$, mRNA degradation rate $\gamma$, and mean burst size
$b$. For our derivation of the chemical master equation we will focus more on
Figure~\ref{figS1:bursty_one_state}(B). This representation is intended to
highlight that bursty gene expression allows transitions between mRNA count $m$
and $m'$ even with $m - m' > 1$.

\begin{figure}[h!]
\centering
\includegraphics{../figures/si/figS0X_bursty_states.pdf}
\caption{
\textbf{Bursty transcription for unregulated promoter.}
(A) Schematic of the one-state bursty transcription model. Rate $k_i$ is the
bursty initiation rate, $\gamma$ is the mRNA degradation rate, and $b$ is the
mean burst size. (B) Schematic depiction of the mRNA count state transitions.
The model in (A) allows for transitions of $> 1$ mRNA counts with probability
$G_{m-m'}$, where the state jumps from having $m'$ mRNA to having $m$ mRNA in a
single burst of gene expression.}
\label{figS1:bursty_one_state}
\end{figure}

To derive the master equation we begin by considering the possible state
transitions to ``enter'' state $m$. There are two possible paths to jump from an
mRNA count $m' \neq m$ to a state $m$ in a small time window $\Delta t$:
\begin{enumerate}
        \item By degradation of a single mRNA, jumping from $m+1$ to $m$.
        \item By producing $m-m'$ mRNA for $m' \in \{0, 1, \ldots, m-1\}$.
\end{enumerate}
For the ``exit'' states from $m$ into $m' \neq m$ during a small time window
$\Delta t$ we also have two possibilities:
\begin{enumerate}
        \item By degradation of a single mRNA, jumping from $m$ to $m-1$.
        \item By producing $m'-m$ mRNA for $m'-m \in \{1, 2, \ldots\}$.
\end{enumerate}
This implies that the probability of having $m$ mRNA at time $t + \Delta t$ can
be written as
\begin{equation}
\begin{split}
p(m, t + \Delta t) = &p(m, t)
+ \overbrace{\gamma \Delta t (m + 1) p(m + 1, t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma \Delta t m p(m, t)}^{m \rightarrow m - 1} \\
&+ \overbrace{k_i \Delta t \sum_{m'=0}^{m-1} G_{m-m'} p(m', t)}^
{m'\in \{0, 1, \ldots m-1\} \rightarrow m}
- \overbrace{k_i \Delta t \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t)}^
{m \rightarrow m'\in \{m+1, m+2, \ldots\}},
\end{split}
\label{eq:si_master_deltat}
\end{equation}
where we indicate $G_{m'-m}$ as the probability of having a burst of size
$m'-m$, i.e. when the number of mRNAs jump from $m$ to $m' > m$ due to a single
mRNA transcription burst. We suggestively use the letter $G$ as we will assume
that these bursts sizes are geometrically distributed with parameter $\theta$.
This is written as
\begin{equation}
G_{k} = \theta (1 - \theta)^k\; \text{for } k \in \{0, 1, 2, \ldots \}.
\end{equation}
In Section~\ref{sec:beyond_means} of the main text we derive this functional
form for the burst size distribution. An intuitive way to think about it is that
for transcription initiation events that take place instantaneously there are
two competing possibilities: Producing another mRNA with probability $(1 -
\theta)$, or ending the burst with probability $\theta$. What this implies is
that for a geometrically distributed burst size we have a mean burst size $b$ of
the form
\begin{equation}
b \equiv \left\langle m' - m \right\rangle 
= \sum_{k=0}^\infty k \theta (1 - \theta)^k = {1 - \theta \over \theta}.
\end{equation}

To clean up Equation~\ref{eq:si_master_deltat} we can send the first term on the
right hand side to the left, and divide both sides by $\Delta t$. Upon taking
the limit where $\Delta t \rightarrow 0$ we can write
\begin{equation}
{d \over dt}p(m, t) = (m + 1) \gamma p(m + 1, t)
- m \gamma p(m, t)
+ k_i \sum_{m'=0}^{m-1} G_{m-m'} p(m', t) 
- k_i \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t).
\end{equation}
Furthermore, given that the timescale for this equation is set by the mRNA
degradation rate $\gamma$ we can divide both sides by this rate, obtaining
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau),
\label{eq:si_master_ode}
\end{equation}
where we defined $\tau \equiv t \times \gamma$, and $\lambda \equiv k_i/\gamma$.
The last term in Eq.~\ref{eq:si_master_ode} sums all burst sizes except for a 
burst of size zero. We can re-index the sum to include this term, obtaining
\begin{equation}
\lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau) = \lambda p(m, t) \left[
\underbrace{\sum_{m'=\textcolor{red}{m}}^{\infty}G_{m'-m}}
_{\text{re-index sum to include burst size zero}} -
\underbrace{G_0}_{\text{subtract extra added term}}\right].
\end{equation}
Given the normalization constraint of the geometric distribution, adding the
probability of all possible burst sizes -- including size zero since we
re-indexed the sum -- allows us to write
\begin{equation}
\sum_{m'=m}^{\infty}G_{m'-m} - G_0 = 1 - G_0.
\end{equation}
Substituting this into Eq.~\ref{eq:si_master_ode} results in
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda p(m, \tau) \left[ 1 - G_0 \right].
\label{eq:si_master_ode_2}
\end{equation}
To finally get at a more compact version of the equation notice that the third
term in Eq.~\ref{eq:si_master_ode_2} includes burst from size $m'-m = 1$ to size
$m' - m = m$. We can include the term $p(m, t) G_0$ in the sum which allows
bursts of size $m' - m = 0$. This results in our final form for the chemical
master equation
\begin{align}
{d \over d\tau}p(m, \tau) = 
(m + 1) p(m+1, \tau)
- m p(m, \tau) - 
\lambda p(m, \tau)
+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime} p(m^\prime, \tau).
\label{eq:si_master_unreg}
\end{align}

In order to solve Eq.~\ref{eq:si_master_unreg} we will use the generating
function method \mrm{reference?}. The probability generating function is defined
as
\begin{align}
F(z,t) = \sum_{m=0}^\infty z^m p(m,t),
\end{align}
where $z$ is just a dummy variable that will help us later on to obtain the
moments of the distribution. Let us now multiply both sides of
Eq.~\ref{eq:si_master_unreg} by $z^m$ and sum over all $m$
\begin{equation}
\sum_m z^m {d \over d\tau} p(m, \tau) = 
\sum_m z^m \left[ 
- m p(m, \tau) 
+ (m + 1) p(m + 1, \tau) 
+ \lambda \sum_{m' = 0}^m G_{m-m'} p(m', \tau) - \lambda p(m, \tau)
\right],
\end{equation}
where we use $\sum_m \equiv \sum_{m=0}^\infty$. We can distribute the sum and
use the definition of $F(z, t)$ to obtain
\begin{equation}
{d F(z, \tau) \over d\tau} =
- \sum_m z^m m p(m, \tau) +
+ \sum_m z^m (m + 1) p(m + 1, \tau)
+ \lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau)
- \lambda F(z, \tau).
\label{eq:si_generating_01}
\end{equation}
We can make use of properties of the generating function to write everything in
terms of $F(z, \tau)$: the first term on the right hand side of
Eq.~\ref{eq:si_generating_01} can be rewritten as
\begin{align}
\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=
\sum_{m} z \frac{\partial z^{m}}{\partial z} p(m, \tau), \\
&=\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
&=z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
&= z {\partial F(z, \tau) \over \partial z}.
\end{align}
For the second term on the right hand side of Eq.~\ref{eq:si_generating_01} we
define $k \equiv m + 1$. This allows us to write
\begin{align}
\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=
\sum_{k=1}^{\infty} z^{k-1} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
&=\frac{\partial F(z)}{\partial z}.
\end{align}

\begin{figure}[h!]
\centering
\includegraphics{../figures/si/figS0X_reindex_sum.pdf}
\caption{\textbf{Reindexing double sum.} Schematic for how to proceed with the
reindexing of the sum $\sum_{m=0}^\infty \sum_{m'=0}^m$. The trick consists on
noticing that this double sum adds up all of the $(m, m')$ pairs in the lower
triangle part of the $m, m'$ plane.}
\label{figS2:sum_reindex}
\end{figure}

The third term in Eq.~\ref{eq:si_generating_01} is the most trouble. The trick
is to reverse the default order of the sums as
\begin{equation}
\sum_{m=0}^{\infty} \sum_{m'=0}^{m} = \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty}.
\end{equation}
To see the logic of the sum we point the reader to
Figure~\ref{figS2:sum_reindex}. The key is to notice that the double sum
$\sum_{m=0}^\infty \sum_{m'=0}^m$ is adding all possible pairs $(m, m')$ in the
lower triangle, so we can add the terms vertically as the original sum
indexing suggests, i.e.
\begin{equation}
\sum_{m=0}^{\infty} \sum_{m'=0}^{m} x_{(m, m')}= 
x_{(0, 0)} + x_{(1, 0)} + x_{(1, 1)} + x_{(2, 0)} + x_{(2, 1)} + x_{(2, 2)} + 
\ldots,
\end{equation}
where the variable $x$ is just a placeholder to indicate the order in which the
sum is taking place. But we can also add the terms horizontally as
\begin{equation}
\sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} x_{(m, m')} =
x_{(0, 0)} + x_{(1, 0)} + x_{(2, 0)} + \ldots + x_{(1,1)} + x_{(2, 1)} + \ldots,
\end{equation}
which still adds all of the lower triangle terms. Applying this reindexing
results in
\begin{align}
\lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau) =
\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau),
\end{align}
where we also substituted the definition of the geometric distribution $G_{k} =
\theta (1 - \theta)^k$. Redistributing the sums we can write
\begin{align}
\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau) = 
\lambda \theta \sum_{n=0}^{\infty}(1-\theta)^{m'} P(m', \tau) 
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m}.
\label{eq:si_generating_02}
\end{align}

The next step requires us to look slightly ahead into what we expect to obtain.
We are working on deriving an equation for the generating function $F(z, \tau)$
that when solved will allow us to compute what we care about, i.e. the
probability function $p(m, \tau)$. Upon finding the function for $F(z, \tau)$,
we will recover this probability distribution by evaluating $F(z=1, \tau)$ in
other cases we can evaluate $F(z=0, \tau)$ to recover instead the moments of the
distribution. The point here is that we know that when all the dust settles we
will evaluate $z$ to be less than one \mrm{someone needs to fact check me}.
Furthermore, we know that the parameter of the geometric distribution $\theta$
must be between zero and one. With these two facts we can safely state that $| z
(1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in
Eq.~\ref{eq:si_generating_02} as
\begin{align}
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &= 
\sum_{n=0}^{\infty} \left[z (1-\theta)\right]^{n + m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{n} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\left( {1 \over 1 - z (1 - \theta)} \right),
\end{align}
where we use the geometric series since, as stated before, $| z (1 - \theta) | <
1$. Putting these results together, the PDE for the generation function is
\begin{equation}
{\partial F \over \partial \tau} = 
{\partial F \over\partial z}
- z {\partial F \over \partial z} - \lambda F
+ \frac{\lambda\theta F}{1-z(1-\theta)}.
\end{equation}
Changing variables to $\xi=1-\theta$ and simplifying gives
\begin{align}
{\partial F \over \partial \tau} + (z - 1) {\partial F \over \partial z} = 
\frac{(z-1)\xi}{1-z\xi}\lambda F.
\label{eq:1state_unreg_015}
\end{align}

\subsubsection{Steady-state}

To get at the mRNA distribution at steady state we first must solve
Eq.~\ref{eq:1state_unreg_015} setting the time derivative to zero. At
steady-state, the PDE reduces to the ODE
\begin{align}
\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
\end{align}
which we can integrate as
\begin{align}
\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
\end{align}
The initial conditions for generating functions can be subtle and confusing. The
key fact follows from the definition
$F(z,t) = \sum_m z^m p(m,t)$,
normalization requires that
$F(z=1, t) = \sum_m p(m,t) = 1$.
\footnote{
Sometimes the generating function may be undefined \textit{at} $z=1$ but the
limit still holds. Also many authors change variables from $z$ to other
things, so one must keep track of how the normalization condition transforms.
}
Doing the integrals (and producing constant $c$) gives
\begin{align}
\ln F &= -\lambda \ln(1-\xi z) + c
\\
F &= \frac{c}{(1-\xi z)^\lambda}.
\end{align}
Only one choice for $c$ can satisfy initial conditions, producing
\begin{align}
F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
\label{eq:gen_fn}
\end{align}

\subsubsection{Recovering the steady-state probability distribution}

To obtain the steady state mRNA distribution $p(m)$ we are aiming for we need to
extract it from the generating function
\begin{equation}
F(z) = \sum_m z^m p(m).
\end{equation}
Taking a derivative with respect to $z$ results in
\begin{equation}
{d F(z) \over dz} = \sum_m m z^{m - 1} p(m).
\end{equation}
Setting $z = 0$ leaves one term in the sum when $m = 1$ 
\begin{equation}
\left.\frac{d F(z)}{d z}\right|_{z=0} = 
\left(0 \cdot 0^{-1} \cdot p(0) 
+ 1 \cdot 0^0 \cdot p(1) 
+ 2 \cdot 0^1 \cdot p(2)
+ \cdots \right) = p(1),
\end{equation}
since in the limit $\lim_{x \rightarrow 0^+} x^x = 1$. A second derivative of
the generating function would result in
\begin{equation}
\frac{d^{2} F(z)}{d z^{2}} = \sum_{m=0}^{\infty} m(m-1) z^{m-2} p(m).
\end{equation}
Again evaluating at $z = 0$ gives 
\begin{equation}
\left.\frac{d^{2} F(z)}{d z}\right|_{z=0} = 2 p(z).
\end{equation}
In general any $p(m)$ is obtained from the generating function as
\begin{equation}
p(m) = {1 \over m!} \left. {d^m F(z) \over dz} \right\vert_{z=0}.
\label{eq:prob_from_gen}
\end{equation}

Let's now look at the general form of the derivative for our generating function
in Eq.~\ref{eq:gen_fn}. For $p(0)$ we simply evaluate $F(z=0)$ directly, 
obtaining
\begin{equation}
p(0) = F(z=0) = \theta^{\lambda}.
\end{equation}
The first derivative results in
\begin{equation}
\begin{aligned}
\frac{d F(z)}{d z} &=\theta^{\lambda} \frac{d}{d z}(1-z(1-\theta))^{-\lambda} \\
&=\theta^{\lambda}\left[-\lambda(1-z(1-f))^{-\lambda-1} \cdot(\theta-1)\right]\\
&=\theta^{\lambda}\left[\lambda(1-z(1-\theta))^{-\lambda-1}(1-\theta)\right].
\end{aligned}
\end{equation}
Evaluating this at $z=0$ as required to get $p(1)$ gives
\begin{equation}
\left.\frac{d F(z)}{d z}\right|_{z=0}=\theta^{\lambda} \lambda(1-\theta)
\end{equation}
For the second derivative we find
\begin{equation}
\frac{d^{2} F(z)}{d z^{2}} = \theta^{\lambda}
\left[\lambda(\lambda+1)(1-z(1-\theta))^{-\lambda-2}(1-\theta)^{2}\right].
\end{equation}
Again evaluating $z = 0$ gives
\begin{equation}
\left.\frac{d^{2} F(z)}{d z^{2}}\right|_{z=0} = 
\theta^{\lambda} \lambda(\lambda+1)(1-\theta)^{2}.
\end{equation}
Let's go for one more derivative to see the pattern. The third derivative of the
generating function gives
\begin{equation}
\frac{d^{3} F(z)}{d z^{3}} = 
\theta^{\lambda} 
\left[\lambda(\lambda+1)
(\lambda+2)(1-z(1-\theta))^{-\lambda-3}(1-\theta)^{3}\right],
\end{equation}
which again we evaluate at $z=0$
\begin{equation}
\left.\frac{d^{3} F(z)}{d z^{3}}\right|_{z=1} =
\theta^{\lambda}\left[\lambda(\lambda+1)(\lambda+2)(1-\theta)^{3}\right].
\end{equation}
If $\lambda$ was an integer we could write this as
\begin{equation}
\left.\frac{d^{3} F(z)}{d z^{3}}\right|_{z=0} = 
\frac{(\lambda+2) !}{(\lambda-1) !} \theta^{\lambda}(1-\theta)^{3}.
\end{equation}
Since $\lambda$ might not be an integer we can write this using Gamma functions
as
\begin{equation}
\left.\frac{d^{3} F(z)}{d z^{3}}\right|_{z=0} = 
\frac{\Gamma(\lambda+3)}{\Gamma(\lambda)} \theta^{\lambda}(1-\theta)^{3}.
\end{equation}
Generalizing the pattern we then have that the $m$-th derivative takes the form
\begin{equation}
\left.\frac{d^{m} F(z)}{d z^{m}}\right|_{z=0} =
\frac{\Gamma(\lambda+m)}{\Gamma(\lambda)} \theta^{\lambda}(1-\theta)^{m}.
\end{equation}
With this result we can use Eq.~\ref{eq:prob_from_gen} to obtain the desired
steady-state probability distribution function
\begin{equation}
p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
        \theta^\lambda (1-\theta)^m.
\end{equation}
Note that the ratio of gamma functions is often expressed as a binomial
coefficient, but since $\lambda$ may be non-integer, this would be ill-defined.
Re-expressing this exclusively in our variables of interest, burst rate
$\lambda$ and mean burst size $b$, we have
\begin{equation}
p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
        \left(\frac{1}{1+b}\right)^\lambda
        \left(\frac{b}{1+b}\right)^m.
\label{eq:nbinom_deriv_final}
\end{equation}
